The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-09-23 paper.qmd
edited for susan comment 11

A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
These studies frequently expect participants to give deterministic answers for probabilistic questions. 
For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and asked them "how likely is candidate B to win the election?". 
Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B". 
The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  
In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". 
The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. 
<!-- It's not unreasonable to use forms of display that are used in gov't communications, even if they're admittedly not optimal -- it can be reasonable to *show* that they're not optimal, and Padilla is usually advocating for different risk displays as a result, including spaghetti maps for the track and probabilistic wind speed/flood risk maps for the impacts. The problem is that gulf coast residents are used to the bad version and not used to interpreting the other versions, so there's a ton of resistance... and I say this as someone who grew up plotting hurricane paths on maps we got from the grocery store in May at the beginning of hurricane season :). These displays are also complicated by the asymmetry of wind/flood risk around the eye of the storm - the northwest side gets hit harder in the northern hemisphere, which leads to natives referring to the "dirty side" and the "clean side" of the storm. Land topology also shapes flood risk, which is yet another complexity. All of this is to say that the explicitly calculated risk maps do a lot of external cognition, but when you're used to doing that calculus mentally, it can be a bit hard to trust the predigested information. -->
Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
Interestingly, it is unclear how the participants were supposed to answer this question. 
@Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth. 
In the first study, participants were required assume to the storm was equally intense regardless of the  probability of the oil rig being hit (i.e. more likely to be hit *does not* mean more damage), but in the second study required participants to assume the opposite (i.e. more likely to be hit *does* mean more damage). 
Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.

>>>>

A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
These studies frequently expect participants to give deterministic answers for probabilistic questions. 
For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and asked them "how likely is candidate B to win the election?". 
Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B". 
The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  
In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". 
The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. 
Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
@Padilla2017 seem to be aware of the issues in the visualisation and are using the user study to show that current government communications are not the best choice when communicating hurricane risk, as it is unreasonable to assume that "more likely to be hit" should automatically translate to "more damage". 
While this is useful to highlight, the authors end up caught by their own trap, as answering the questions in their second study correctly required participants to assume an area being more likely to be hit *does* translate to receiving more damage.
The assumptions required to answer these deterministic questions are often invisible, even to the authors of the paper, which means they often create more problems than they solve when they are used as a tool for graphical evaluation.
Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.
